Division and Factorization of p-Adic Polynomials

### 2016-01-05 polynomials/muldivfact.tex
Draft for the proof of the slope factorization theorem

\begin{deftn}
Let $\varphi$ and $\psi$ be two Newton functions of degree $n$ and
$d$ respectively. Letting $\mu = \psi(d) - \psi(d-1)$ and $\Delta$
be the greatest affine function of slope $\mu$ with $\Delta \leq
\varphi$, we define:
$$\begin{array}{rcl}
\varphi \nfmod \psi \, : \, [0, d-1] & \to & \R \cup \{+\infty\} \smallskip \\
x & \mapsto & \psi(x) + \Delta(d) - \psi(d) \medskip \\
\varphi \nfdiv \psi \, : \, [0, n-d] & \to & \R \cup \{+\infty\} \smallskip \\
x & \mapsto & \inf_{h \geq 0} \psi(x+d+h) - \mu h.
\end{array}$$
\end{deftn}

@@

Figure~\ref{fig:NewtonEuclide} illustrates the definition: if $\varphi$ 
and $\psi$ are the functions represented on the diagram, the epigraph of 
$\varphi \nfmod \psi$ is the blue area whereas that of $\varphi \nfdiv \psi$ 
is the green area translated by $(-d,0)$.
It is an easy exercise (left to the reader) to design algorithms for
computing $\varphi \nfmod \psi$ and $\varphi \nfdiv \psi$ for a cost
which is at most $n + O(1)$.

@@

\begin{rem}
Keeping the notations of the Theorem and decomposing $A$ as a sum 
$A_{<d} + A_{\geq d}$ where $A_{<d}$ (resp. $A_{\geq d}$) consists of
monomials of $A$ of degree $< d$ (resp of degree $\geq d$), we have:
$$A \div B = A_{\geq d} \div B 
\quad \text{and} \quad
A \mod B = A_{<d} + (A_{\geq d} \mod B).$$
Using the above relations and applying Theorem~\ref{theo:EDivisionNP}
with $A_{\geq d}$ and $B$ (instead of $A$ and $B$ respectively) may lead
to a better lower bound on the Newton function of $A \div B$ and $A \mod
B$.
\end{rem}

\begin{proof}[of Theorem~\ref{theo:EDivisionNP}]
Let us first prove Eq.~\eqref{eq:AmodB}. 
Replacing $B$ by $c^{-1} B$ where $c$ denotes the leading coefficient
of $B$, we may assume that $B$ is monic. Using linearity, we may further
assume that $A$ is a monomial. Set $R_n = X^n \mod B$. The relation
we have to prove is:
$$\NF(R_n)(x) \geq \NF(B)(x) - \mu(n-d).$$
When $n < d$, we have $R_n = 0$ and the assertion of the Theorem holds. 
For larger $n$, we proceed by induction. We have the relation
$R_{n+1} = X R_n - c_n B$
where $c_n$ is the coefficient in $X^{d-1}$ of $R_n$. Thanks to the
induction hypothesis, we have:
\begin{align*}
\val(c_n) \geq \NF(R_n)(d{-}1) & \geq \NF(B)(d{-}1) - \mu(n{-}d) \\
& = - \mu(n{+}1{-}d)
\end{align*}
since $\mu = -\NF(B)(d{-}1)$ because $B$ is monic. Therefore
$\NF(c_n B)(x) \geq \NF(B)(x) - \mu(n+1-d)$ for all $x$. On the 
other hand, for all $x$, we have:
$$\NF(X R_n)(x) = \NF(R_n)(x{-}1) \geq \NF(R_n)(x) - \mu$$
from what we get $\NF(X R_n)(x) \geq \NF(B)(x) - \mu(n+1-d)$. As a
consequence
$\NF(R_{n+1})(x) \geq \NF(B)(x) - \mu(n+1-d)$
and the induction follows.
Eq.~\eqref{eq:AdivB} is now derived from:
$$\NF(A \div B) \nftimes \NF(B) \geq \NF(A) \nfplus \NF(A \mod B)$$
using the estimation on $\NF(A\mod B)$ we have just proved
(see Figure~\ref{fig:NewtonEuclide}).
\end{proof}

@@

\begin{proof}
Let $\delta A$ (resp. $\delta B$) be a polynomial whose Newton 
function is not less than $\varphi_A$ (resp. $\varphi_B$) and define
$\delta Q$ and $\delta R$ by:
\begin{align*}
Q_0 + \delta Q & = (A_0 + \delta A) \div (B_0 + \delta B) \\
R_0 + \delta R & = (A_0 + \delta A) \mod (B_0 + \delta B)
\end{align*}
where $Q_0 = A_0 \div B_0$ and $R_0 = A_0 \mod B_0$.
We have to show that $\NF(\delta Q) \geq \varphi \nfdiv \NF(B)$ and
$\NF(\delta R) \geq \varphi \nfmod \NF(B)$.
Set $\delta X = \delta A - Q_0 \delta B$. 
Using Theorem~\ref{theo:EDivisionNP}, we obtain $\NF(Q_0) \geq
\NF(A) \div \NF(B)$ and consequently $\NF(\delta X) \geq \varphi$.
On the other hand, an easy computation yields
$$\delta X = (B + \delta B) \cdot \delta Q + \delta R$$
so that $\delta Q = \delta X \div (B + \delta B)$ and 
$\delta R = \delta X \mod (B + \delta B)$. Using again 
Theorem~\ref{theo:EDivisionNP}, we get the desired result.
\end{proof}

@@

\begin{theo} \label{theo:slope-factor}
Let $P(X) = \sum_{i=0}^n a_i X^i$ be a polynomial of degree $n$ with
coefficients in $K$. We assume that $\NP(P)$ has an extremal point 
whose abscissa is $d$.
We define the sequences $(A_i)_{i \geq 0}$ and $(V_i)_{i \geq 0}$
recursively by:
\begin{align*}
A_0 & = \sum_{i=0}^d a_i X^i, \quad V_{0} = 1 \medskip \\
A_{i+1} &= A_i + (V_i P \:\mod\: A_i), \smallskip \\
V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \mod A_{i+1} \\
& \hspace{2cm}\text{where } B_{i+1} = P \div A_{i+1}.
\end{align*}
Then the sequence $(A_i)$ convergences to a divisor $A$ of $P$ 
of degree $d$ whose leading coefficient is $a_d$ and whose Newton 
function agrees with $\NF(P)$ on $[0,d]$.
\end{theo}

@@

\begin{lem} \label{lem:formulae}
The following relations hold:
\begin{align}
\Delta B_i &= -(R_i B_{i+1}) \div A_i, \label{eqdef:Biminus} \\
\Delta S_i &= -(R_i B_{i+1}) \mod A_i,  \label{eqdef:Siminus} \\
S_{i} &= (B_i R_i + T_i S_{i-1} - T_i \: \Delta S_{i-1}) \mod A_{i}, \label{eqdef:Si2}  \\
\Delta V_i &= (V_i T_i - V_i^2 \: \Delta B_i) \mod A_{i}, \label{eqdef:Viminus}\\
1-Q_i &= T_i + (R_i-V_i S_i) \div A_i, \label{eqdef:Qiminus}\\
R_{i+1} &= (\Delta V_i \: S_{i+1}+(1{-}Q_i)R_i) \mod A_{i+1}, \label{eq:Riplus}\\
T_{i+1} &\equiv T_i + V_i \: (\Delta B_i)^2 \pmod{A_{i+1}}. \label{eq:ViBiplus}
\end{align}
\end{lem}

@@

We have $V_i P=A_i Q_i+R_i=(A_{i+1}-R_i)Q_i+R_i$ and $V_{i+1} P=A_{i+1} 
Q_{i+1}+R_{i+1}$. Thus:
\begin{align*}
R_{i+1} &= \Delta V_i \: P + (1-Q_i)R_i \\
 &= (\Delta V_i \: S_{i+1} + (1-Q_i)R_i) \mod A_{i+1},
\end{align*} 
and \eqref{eq:Riplus} is proved. Finally
\begin{align*}
T_{i+1} &\equiv 1-2V_i B_{i+1}+V_i^2 B_{i+1}^2 \pmod {A_{i+1}} \\
&\equiv (1-V_i B_{i+1})^2 \pmod {A_{i+1}} \\
&\equiv T_i + V_i (\Delta B_i)^2 \pmod {A_{i+1}}
\end{align*}
which concludes the proof.
\end{proof}

>>>>

\begin{deftn}
Let $\varphi$ and $\psi$ be two Newton functions of degree $n$ and
$d$ respectively. Set $\mu = \psi(d) - \psi(d-1)$. Letting $\Delta$
be the greatest affine function of slope $\mu$ with $\Delta \leq
\varphi_{|[d,n]}$ and $\delta = \Delta(d) - \psi(d)$, we define:
$$\begin{array}{r@{\,\,}rcl}
\varphi \nfmod \psi & 
\multicolumn{3}{@{}l}{= \varphi_{|[0,d{-}1]} \nfplus \big( 
\psi_{|[0,d{-}1]} + \delta\big)} \medskip \\
\varphi \nfdiv \psi & : \, [0, n-d] & \to & \R \cup \{+\infty\} \smallskip \\
& x & \mapsto & \inf_{h \geq 0} \psi(x+d+h) - \mu h.
\end{array}$$
\end{deftn}

@@

Figure~\ref{fig:NewtonEuclide} illustrates the definition: if $\varphi$ 
and $\psi$ are the functions represented on the diagram, the epigraph of 
$\varphi \nfmod \psi$ is the blue area whereas that of $\varphi \nfdiv \psi$ 
is the green area translated by $(-d,0)$.
It is an easy exercise (left to the reader) to design quasi-linear
algorithms for computing $\varphi \nfmod \psi$ and $\varphi \nfdiv \psi$.

@@

\begin{proof}
Write $A = A_{<d} + A_{\geq d}$ where $A_{<d}$ (resp. $A_{\geq d}$)
consists of monomials of $A$ of degree at least $d$ (resp. less than
$d$). Noting that:
$$A \mod B = A_{<d} + (A_{\geq d} \mod B)
\quad \text{and} \quad
A \div B = A_{\geq d} \div B$$
we may assume that $A = A_{\geq d}$.

Let us now prove Eq.~\eqref{eq:AmodB}. 
Replacing $B$ by $c^{-1} B$ where $c$ denotes the leading coefficient
of $B$, we may assume that $B$ is monic. Using linearity, we may further
assume that $A$ is a monomial. Set $R_n = X^n \mod B$. The relation
we have to prove is:
$$\NF(R_n)(x) \geq \NF(B)(x) - \mu(n-d)
\quad \text{for } n\geq d.$$
We proceed by induction. The initialisation is clear because $R_d$
agrees with $(-B)$ up to degree $d{-}1$. We have the relation
$R_{n+1} = X R_n - c_n B$
where $c_n$ is the coefficient in $X^{d-1}$ of $R_n$. Thanks to the
induction hypothesis, we have:
\begin{align*}
\val(c_n) \geq \NF(R_n)(d{-}1) & \geq \NF(B)(d{-}1) - \mu(n{-}d) \\
& = - \mu(n{+}1{-}d)
\end{align*}
since $\mu = -\NF(B)(d{-}1)$ because $B$ is monic. Therefore
$\NF(c_n B)(x) \geq \NF(B)(x) - \mu(n{+}1{-}d)$ for all $x$. On the 
other hand, for all $x$, we have:
$$\NF(X R_n)(x) = \NF(R_n)(x{-}1) \geq \NF(R_n)(x) - \mu$$
from what we get $\NF(X R_n)(x) \geq \NF(B)(x) - \mu(n{+}1{-}d)$. As a
consequence
$\NF(R_{n+1})(x) \geq \NF(B)(x) - \mu(n{+}1{-}d)$
and the induction follows.
Eq.~\eqref{eq:AdivB} is now derived from:
$$\NF(A \div B) \nftimes \NF(B) \geq \NF(A) \nfplus \NF(A \mod B)$$
using the estimation on $\NF(A\mod B)$ we have just proved
(see Figure~\ref{fig:NewtonEuclide}).
\end{proof}

@@

\begin{proof}
Let $\delta A$ (resp. $\delta B$) be a polynomial whose Newton 
function is not less than $\varphi_A$ (resp. $\varphi_B$) and define
$\delta Q$ and $\delta R$ by:
\begin{align*}
Q_0 + \delta Q & = (A_0 + \delta A) \div (B_0 + \delta B) \\
R_0 + \delta R & = (A_0 + \delta A) \mod (B_0 + \delta B)
\end{align*}
where $Q_0 = A_0 \div B_0$ and $R_0 = A_0 \mod B_0$.
We have to show that $\NF(\delta Q) \geq \varphi \nfdiv \NF(B)$ and
$\NF(\delta R) \geq \varphi \nfmod \NF(B)$.
Set $\delta X = \delta A - Q_0 \delta B$. 
Using Theorem~\ref{theo:EDivisionNP}, we obtain $\NF(Q_0) \geq
\NF(A) \nfdiv \NF(B)$ and consequently $\NF(\delta X) \geq \varphi$.
On the other hand, an easy computation yields
$$\delta X = (B + \delta B) \cdot \delta Q + \delta R$$
so that $\delta Q = \delta X \div (B + \delta B)$ and 
$\delta R = \delta X \mod (B + \delta B)$. Using again 
Theorem~\ref{theo:EDivisionNP}, we get the desired result.
\end{proof}

@@

\begin{theo} \label{theo:slope-factor}
Let $P(X) = \sum_{i=0}^n a_i X^i$ be a polynomial of degree $n$ with
coefficients in $K$. We assume that $\NP(P)$ has an extremal point 
whose abscissa is $d$.
We define the sequences $(A_i)_{i \geq 0}$ and $(V_i)_{i \geq 0}$
recursively by:
\begin{align*}
A_0 & = \sum_{i=0}^d a_i X^i, \quad V_{0} = 1 \medskip \\
A_{i+1} &= A_i + (V_i P \:\mod\: A_i), \smallskip \\
V_{i+1} &= (2 V_i -V_i^2 B_{i+1} ) \mod A_{i+1} \\
& \hspace{2cm}\text{where } B_{i+1} = P \div A_{i+1}.
\end{align*}
Then the sequence $(A_i)$ converges to a divisor $A_\infty$ of $P$ 
of degree $d$ whose leading coefficient is $a_d$ and whose Newton 
function agrees with $\NF(P)$ on $[0,d]$.
Moreover, setting:
$$\kappa = \frac{\NF(P)(d{+}1) + \NF(P)(d{-}1)} 2 - \NF(P)(d)$$
(with $\NF(P)(-1) = \NF(P)(n{+}1) = +\infty$ if necessary),
we have $\kappa > 0$ and the following rate of convergence:
$$\forall i \geq 1, \quad
\NF(A_\infty - A_i) \geq \NF(P)_{|[0,d]} \,+\, 2^i \kappa.$$
\end{theo}

@@

\begin{lem} \label{lem:formulae}
The following relations hold:
\begin{align}
\Delta B_i &= -(R_i B_{i+1}) \div A_i, \label{eqdef:Biminus} \\
\Delta S_i &= -(R_i B_{i+1}) \mod A_i,  \label{eqdef:Siminus} \\
S_{i} &= (B_i R_i + T_i S_{i-1} - T_i \: \Delta S_{i-1}) \mod A_{i}, \label{eqdef:Si2}  \\
\Delta V_i &= (V_i T_i - V_i^2 \: \Delta B_i) \mod A_{i}, \label{eqdef:Viminus}\\
1-Q_i &= T_i + (R_i-V_i S_i) \div A_i, \label{eqdef:Qiminus}\\
R_{i+1} &= (\Delta V_i \: S_{i+1}+(1{-}Q_i)R_i) \mod A_{i+1}, \label{eq:Riplus}\\
T_{i+1} &\equiv (T_i + V_i \: \Delta B_i)^2 \pmod{A_{i+1}}. \label{eq:ViBiplus}
\end{align}
\end{lem}

@@

We have $V_i P=A_i Q_i+R_i=(A_{i+1}-R_i)Q_i+R_i$ and $V_{i+1} P=A_{i+1} 
Q_{i+1}+R_{i+1}$. Thus:
\begin{align*}
R_{i+1} &= \Delta V_i \: P + (1-Q_i)R_i \\
 &= (\Delta V_i \: S_{i+1} + (1-Q_i)R_i) \mod A_{i+1},
\end{align*} 
and \eqref{eq:Riplus} is proved. Finally
\begin{align*}
T_{i+1} &\equiv 1-2V_i B_{i+1}+V_i^2 B_{i+1}^2 \pmod {A_{i+1}} \\
&\equiv (1-V_i B_{i+1})^2 \pmod {A_{i+1}} \\
&\equiv (T_i + V_i \: \Delta B_i)^2 \pmod {A_{i+1}}
\end{align*}
which concludes the proof.
\end{proof}
